---
title: "Numerical MLE of Gamma Parameters"
subtitle: "Task III. Programming in R. MSc Statistics for Data Science."
author:
  - "Kendal Raymond William Smith <100494805@alumnos.uc3m.es>"
  - "Germán Blanco Blanco <100441287@alumnos.uc3m.es>"
date: '2022-10-22'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
if (!require("maxLik")) install.packages("maxLik")

library(maxLik)
```

# Problem Statement

We are going to compare two different ways to approximate the parameters of a gamma distribution using R libraries.
The first one uses the standard optimization package (optim and optimize functions) and the second one the MaxLik package.
The input data will be a set of 5000 samples generated with the rgamma function.

# Solution Using Standard Optimization Functions in R

The following code generates a list of 5000 samples from a gamma distribution and calculates the log_likelihood for those samples.

```{r}
sample <- rgamma(5000, shape=3, rate=3)

log_likelihood <- function(sample, alpha, beta) {
  return(sum(log(dgamma(sample, shape=alpha, rate=beta))))
}

# This is log likelihood without parameters
# It will be used in the next section

ll <- function(x) {
  alpha <- x[1]
  beta <- x[2]
  return(log_likelihood(sample, alpha, beta))
}

# This is the negative log likelihood without parameters
# so that everything works without any more tweaks in optim

nll <- function(x) {
  return(-ll(x))
}
```

Now we estimate the gamma parameters using the optim function.

```{r}

# Initial values given by the Method of Moments
alpha_0 = mean(sample)^2/var(sample)
beta_0 = alpha_0/mean(sample)

optim(c(alpha_0, beta_0), nll)

```

We can also use initial values that are separated from the real values of the parameters to check that the method works also in this case.

```{r}

# Test with different initial values

optim(c(2, 4), nll)

```

# Solution Using MaxLik Package

The code using the MaxLik package is very similar, however it seems to work faster and with less iterations.

```{r}

# Initial values given by the Method of Moments
maxLik(ll, start = c(alpha=alpha_0, beta=beta_0) )

```

```{r}

# Different initial values
maxLik(ll, start = c(alpha=2, beta=4) )

```

# Conclusion

MaxLik provides a very similar interface to the optim function, but it uses different algorithms (in the example Newton-Raphson) and it seems to have integrated estimation of the gradient (Newton-Raphson requires use of the gradient, which is not provided to the MaxLik function).
